Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-18 projectReport/chapter1.tex
Worked through all tex errors

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually `negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. We encounter many difficultiesin the process of modelling such systems:
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The system may be described by functions that behave wildly because their graphs have a wild oscillatory behaviour, i.e., a large functional complexity.
\end{enumerate}

@@

Examples in the literature attempting to forecast data from such systems have tried many different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (\cite{brunton2016discovering, champion2019data} and \cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

Recent partially interpretable models available in the literature have been based on the Koopman operator (see~\cite{koopman1932dynamical,budivsic2012applied}) to employ observables mapping the data onto a higher-dimensional space. This makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods do not guarantee exact reconstructions for nonlinear models, and in practice provide poor long-term consistency for a large class of chaotic dynamical systems \cite{Supp}.

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding method of delay-coordinates~\cite{takens1981detecting} when one has "good" observations from the system; this learnt system could then be used to forecast the observed data. 
Takens delay embedding theorem~\cite{takens1981detecting} and its various generalisations (see,~\cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the comprehension of the dynamics unreliable (see, [?]). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter~ \ref{ch3}), it can cause predictions to fail.

@@

\emph{By solidifying the mathematical underpinnings of our theory, we hope to guarantee the ability to construct models with predictive power ranging from molecular biology to neuroscience. in the near future. (We can modify this sentence when the report is completed.)}

>>>>

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. 
Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. 
For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. 
However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually `negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. We encounter many difficultiesin the process of modelling such systems:

\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The system may be described by functions that behave wildly because their graphs have a wild oscillatory behaviour, i.e., a large functional complexity.
\end{enumerate}

@@

Examples in the literature attempting to forecast data from such systems have tried many different approaches with differing degrees of success. 
When we have complete access to states of the system, an ordinary differential equation can be obtained from data (\cite{brunton2016discovering, champion2019data} and~\cite{small2002modeling,xu2006modeling}) wherein one could approximate the vector field by a library of functions to obtain interpretable models.

Recent partially interpretable models available in the literature have been based on the Koopman operator (see~\cite{koopman1932dynamical,budivsic2012applied}) to employ observables mapping the data onto a higher-dimensional space. This makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods do not guarantee exact reconstructions for nonlinear models, and in practice provide poor long-term consistency for a large class of chaotic dynamical systems~\cite{Supp}.

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding method of delay-coordinates~\cite{takens1981detecting} when one has `good' observations from the system; this learnt system could then be used to forecast the observed data. 
Takens delay embedding theorem~\cite{takens1981detecting} and its various generalisations (see,~\cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the comprehension of the dynamics unreliable (see, [?]). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter~\ref{ch3}), it can cause predictions to fail.

@@

\emph{By solidifying the mathematical underpinnings of our theory, we hope to guarantee the ability to construct models with predictive power ranging from molecular biology to neuroscience in the near future. (We can modify this sentence when the report is completed.)}

### 2022-01-18 projectReport/chapter2.tex
Worked through all tex errors

This chapter briefly describes a discrete-time dynamical system and what it means to exhibit chaos. We refer to \cite{devaney2018introduction, de2013elements} for more details.

@@

\begin{Definition}
  [\bf Flow of an Equation] \label{Dfn_Flow}\rm
  The flow of the equation  $\dot{x} = f(x)$  is defined to be a mapping $\varphi: \mathbb{R}^n \times \mathbb{R} \to \mathbb{R}$ where $\varphi(x_0,t)= x(t)$ for the solution $x(t)$ with $x(0)=x_0$ 
\end{Definition}

@@

A core concept in the study of dynamical systems is that of invariance. 
\begin{Definition}
  [\bf Invariant Set]\label{Dfn_InvariantSet}\rm
  Given a discrete-time dynamical system $T: U \to U$, a subset $A \subset U$ is said to be an \emph{invariant set} if $T(A) =A$. 
\end{Definition}

We also define the orbit of a function $T$.
\begin{Definition}
  [\bf Orbit of $T$]\label{Dfn_Orbit}\rm
  The orbit of $T$ is to be defined the sequence $\bar{u} = \{u_n\}_{n\in \mathbb{Z}}$ obeying the update equation, $u_{n+1}=Tu_n$, $n \in \mathbb{Z}$. 
\end{Definition}

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^pu\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could be also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant (as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$).

@@

However,  when the space $U$ is compact, all orbits approach an invariant set. This is since the set of limit points of the orbit can be shown to be invariant \cite{de2013elements}. So, when $U$ is compact,  the $\omega$-limit set $\omega(u;T)$ of a point $u$ defined to the collection of limit points of the sequence $\{x,Tu,T^2u,\ldots\}$ is nonempty, and $\omega(u;T)$ is invariant.

\begin{Example}
  For the map $Tu=u^2$ defined on $[0,1]$ all orbits lie in the invariant set $\{1\}$ or else would would approach the invariant set $\{0\}$. 
\end{Example}

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in their vicinity behave. 
Recall the examples $Tu =u/2$ and $Tu=2u$ defined on $\mathbb{R}$ (where $\{0\}$ “attracts" orbits) or the example, $Tu=u^2$ defined on $[0,1]$ (where $\{1\}$ “repels" orbits). 
We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
In particular, we are interested in those invariant sets named attractors.

\begin{Definition}
  [\bf Attractor]\label{Dfn_Attractor}\rm
  Let $T: U \to U$, where $U$  is a metric space with metric $d$. A compact subset $A \subset U$ is said to be an attractor if it satisfies the three conditions: 
  \vspace{-8mm}
  \begin{enumerate}
	\item $A$ is invariant. 
	\item $A$ is asymptotically stable, i.e., for every $\epsilon > 0$ and for all $u$ so that $d(u,A) < \epsilon$, we have $d(T^nu,A) \to 0$ as $n\to \infty$. 
	\item $A$ has Lyapunov stability, i.e., for every $\epsilon > 0$  there exists a $\delta(\epsilon) > 0$ so that $d(u,A) < \delta$ implies $d(T^nu,A) < \epsilon$ for all $n\ge 0$.  
\end{enumerate}
\end{Definition}

@@

In the 1960s, several mathematicians and mathematically interested scientists independently discovered chaos in the mathematical sense. The meteorologist Edward Lorenz may have been the first to explain this phenomenon in his 1963 paper \cite{lorenz1963deterministic}. The notions of invariance, attractivity, and chaos may also be described for continuous systems, and Lorenz's system comprised of a system of differential equations. 
The narrative of Lorenz's discovery of chaos and the history of other forerunners in this subject is fascinating. We highly recommend James Gleick's book Chaos: The Making of a New Science~\cite{gleick2008chaos} for those interested. It clearly illustrates these experiences, explains why chaos was such a startling and crucial mathematical and scientific discovery and describes the underlying mathematical notions for non-specialists.

@@

In practice, when only data is observed from a system, it is not possible to verify which definition or notion is satisfied by the underlying dynamical system. We do, however, specifically recall the definition of chaos in the sense of Devaney \cite{devaney2018introduction,de2013elements} so as to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion  of sensitive dependence on initial conditions.

@@

To illustrate the concept of SDIC, we present two examples - one from Mathematics and the other from the field of Physics.

\begin{Example} \rm
  The logistic map(LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$ is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC. \ednote{B: edited and removed unnecessary citation}
  In the graph below~\ref{fig:log_sdic} we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$.

@@

\begin{Example} \rm
  A second example pertains to a simple physical system called the double pendulum - a pendulum with another pendulum attached to its head given an initial position and angular velocity.

@@

Topological transitivity implies that iterates of an open set of initial conditions get mixed up with other open sets. On a compact metric space, one may show that topological transitivity also implies the existence of a point whose forward iterates are dense \cite{de2013elements}; or in other words, the orbit going through this point will be dense in the compact metric space. 
In fact, it is topologically more likely that the choice of an arbitrary point will be one whose iterates are almost dense.

@@

To be more precise (see \cite{de2013elements}), for a discrete-time dynamical system on a compact space, the set of points with dense iterates are residual, and they are typical or likely to be observed in practice. In this project, when data is observed from a topologically transitive system, we assume it arises from a dense orbit.

@@

  We now reason as to why this is true. The graph of the map $T$ is piecewise linear with two straight lines, one connecting the points $(0,0)$ and $(\frac{1}{2},1)$ and the other connecting $(\frac{1}{2},1)$ with $(1,0)$. They form a so-called tent with base centered at $u=1/2$. The graph of the map $T^2$ comprises two symmetric tents with their base centered at $1/4$ and $3/4$. See Figure \ref{fig:T2tentmap}

@@

%Composite functions
To understand the diagram, we note that if we travel "right and then down," the diagram instructs us to use $T$ (top arrow right) first, followed by $\phi$. (right arrow downwards). Consequently, the pathway amounts to finding $\phi(T(u))$. If we proceed "down and then right," the diagram instructs us to apply $\phi$ (left arrow downwards) first, and then apply $S$ (bottom arrow right). This then amounts to finding  $S(\phi(u))$. When the relationships in this diagram hold, we say $\phi(T(u))= S(\phi(u))$, and we formally denote it as $\phi \circ T=S\circ \phi$.

@@

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see, \cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

It is a very hard or nearly an impossible task to establish the existence of a conjugacy or a semi-conjugacy $\phi$ between two systems \cite{devaney2018introduction}. However, one can verify that a function $\phi$ satisfies the commutativity diagram \ref{eqn_conjugacy}.

\begin{Example}\rm
  For example $\phi(x)=\sin\big(\frac{\pi}{2}x\big)^2$ is a conjugacy between two systems $Tu=1-|2u-1|$ and $Sv=4v(1-v)$  defined on $U=[0,1]$.

  Indeed, $\phi:[0,1]\to[0,1]$ is a homeomorphism and by employing simple trigonometric identities, it is easily proved that $\phi$ satisfies~\ref{eqn_conjugacy}
  \[
    \begin{aligned}[t]
      (\phi\circ{T})u 
                &=\sin^{2}\Big(\frac{\pi}{2} - \frac{\pi}{2}|2u-1|\Big) = \sin^{2}(\pi{u})=\Big(2\sin(\frac{\pi}{2})\cos(\frac{\pi}{2})\Big)^{2}=4\sin^{2}(\frac{\pi}{2}u)\Big(1-\sin^{2}\frac{\pi}{2}u\Big) \\
                                    &=({S}\circ\phi)u.
    \end{aligned}
\]
And it follows that $\phi$ is a conjugacy between the two systems.
\end{Example}

>>>>

This chapter briefly describes a discrete-time dynamical system and what it means to exhibit chaos. We refer to~\cite{devaney2018introduction, de2013elements} for more details.

@@

\begin{Definition}
  [\bf Flow of an Equation]\label{Dfn_Flow}\rm
  The flow of the equation  $\dot{x} = f(x)$  is defined to be a mapping $\varphi: \mathbb{R}^n \times \mathbb{R} \to \mathbb{R}$ where $\varphi(x_0,t)= x(t)$ for the solution $x(t)$ with $x(0)=x_0$ 
\end{Definition}

@@

A core concept in the study of dynamical systems is that of invariance. 
\begin{Definition}
  [\bf Invariant Set]\label{Dfn_InvariantSet}\rm
  Given a discrete-time dynamical system $T: U \to U$, a subset $A \subset U$ is said to be an \emph{invariant set} if $T(A) =A$. 
\end{Definition}
We also define the orbit of a function $T$.
\begin{Definition}
  [\bf Orbit of $T$]\label{Dfn_Orbit}\rm
  The orbit of $T$ is to be defined the sequence $\bar{u} = {\{u_n\}}_{n\in \mathbb{Z}}$ obeying the update equation, $u_{n+1}=Tu_n$, $n \in \mathbb{Z}$. 
\end{Definition}

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^{pu}\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could be also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant (as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$).

@@

However,  when the space $U$ is compact, all orbits approach an invariant set. This is since the set of limit points of the orbit can be shown to be invariant~\cite{de2013elements}. So, when $U$ is compact,  the $\omega$-limit set $\omega(u;T)$ of a point $u$ defined to the collection of limit points of the sequence $\{x,Tu,T^2u,\ldots\}$ is nonempty, and $\omega(u;T)$ is invariant.

\begin{Example}\rm
  For the map $Tu=u^2$ defined on $[0,1]$ all orbits lie in the invariant set $\{1\}$ or else would would approach the invariant set $\{0\}$. 
\end{Example}

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in their vicinity behave. 
Recall the examples $Tu =u/2$ and $Tu=2u$ defined on $\mathbb{R}$ (where $\{0\}$ `attracts' orbits) or the example, $Tu=u^2$ defined on $[0,1]$ (where $\{1\}$ `repels' orbits). 
We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
In particular, we are interested in those invariant sets named attractors.

\begin{Definition}
  [\bf Attractor]\label{Dfn_Attractor}\rm
  Let $T: U \to U$, where $U$  is a metric space with metric $d$. A compact subset $A \subset U$ is said to be an attractor if it satisfies the three conditions: 
  \vspace{-8mm}
  \begin{enumerate}
	\item $A$ is invariant. 
	\item $A$ is asymptotically stable, i.e., for every $\epsilon > 0$ and for all $u$ so that $d(u,A) < \epsilon$, we have $d(T^nu,A){\to}{0}$ as $n\to \infty$. 
	\item $A$ has Lyapunov stability, i.e., for every $\epsilon > 0$  there exists a $\delta(\epsilon) > 0$ so that $d(u,A) < \delta$ implies $d(T^nu,A) < \epsilon$ for all $n\ge 0$.  
\end{enumerate}
\end{Definition}

@@

In the 1960s, several mathematicians and mathematically interested scientists independently discovered chaos in the mathematical sense. The meteorologist Edward Lorenz may have been the first to explain this phenomenon in his 1963 paper~\cite{lorenz1963deterministic}. The notions of invariance, attractivity, and chaos may also be described for continuous systems, and Lorenz's system comprised of a system of differential equations. 
The narrative of Lorenz's discovery of chaos and the history of other forerunners in this subject is fascinating. We highly recommend James Gleick's book Chaos: The Making of a New Science~\cite{gleick2008chaos} for those interested. It clearly illustrates these experiences, explains why chaos was such a startling and crucial mathematical and scientific discovery and describes the underlying mathematical notions for non-specialists.

@@

In practice, when only data is observed from a system, it is not possible to verify which definition or notion is satisfied by the underlying dynamical system. We do, however, specifically recall the definition of chaos in the sense of Devaney~\cite{devaney2018introduction,de2013elements} so as to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion  of sensitive dependence on initial conditions.

@@

To illustrate the concept of SDIC, we present two examples -- one from Mathematics and the other from the field of Physics.

\begin{Example} \rm
  The logistic map (LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$ is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC.
    \ednote{B:edited and removed unnecessary citation}
  In Graph~\ref{fig:log_sdic} below, we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$.

@@

\begin{Example} \rm
  A further example pertains to the simple physical system of the double pendulum -- a pendulum with another pendulum attached to its head given an initial position and angular velocity (see the setup of the pendulum in Figure~\ref{fig:dp_setup}).

@@

Topological transitivity implies that iterates of an open set of initial conditions get mixed up with other open sets. On a compact metric space, one may show that topological transitivity also implies the existence of a point whose forward iterates are dense~\cite{de2013elements}; or in other words, the orbit going through this point will be dense in the compact metric space. 
In fact, it is topologically more likely that the choice of an arbitrary point will be one whose iterates are almost dense.

@@

To be more precise (see~\cite{de2013elements}), for a discrete-time dynamical system on a compact space, the set of points with dense iterates are residual, and they are typical or likely to be observed in practice. In this project, when data is observed from a topologically transitive system, we assume it arises from a dense orbit.

@@

  We now reason as to why this is true. The graph of the map $T$ is piecewise linear with two straight lines, one connecting the points $(0,0)$ and $(\frac{1}{2},1)$ and the other connecting $(\frac{1}{2},1)$ with $(1,0)$. They form a so-called tent with base centered at $u=1/2$. The graph of the map $T^2$ comprises two symmetric tents with their base centered at $1/4$ and $3/4$. See Figure~\ref{fig:T2tentmap}

@@

%Composite functions
To understand the diagram, we note that if we travel `right and then down', the diagram instructs us to use $T$ (top arrow right) first, followed by $\phi$. (right arrow downwards). Consequently, the pathway amounts to finding $\phi(T(u))$. If we proceed `down and then right', the diagram instructs us to apply $\phi$ (left arrow downwards) first, and then apply $S$ (bottom arrow right). This then amounts to finding  $S(\phi(u))$. When the relationships in this diagram hold, we say $\phi(T(u))= S(\phi(u))$, and we formally denote it as $\phi \circ T=S\circ \phi$.

@@

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see,~\cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

It is a very hard or nearly an impossible task to establish the existence of a conjugacy or a semi-conjugacy $\phi$ between two systems~\cite{devaney2018introduction}. However, one can verify that a function $\phi$ satisfies the commutativity diagram~\ref{eqn_conjugacy}.

\begin{Example}\rm
  For example $\phi(x)=\sin{\big(\frac{\pi}{2}x\big)}^2$ is a conjugacy between two systems $Tu=1-|2u-1|$ and $Sv=4v(1-v)$  defined on $U=[0,1]$.

  Indeed, $\phi:[0,1]\to[0,1]$ is a homeomorphism and by employing simple trigonometric identities, it is easily proved that $\phi$ satisfies~\ref{eqn_conjugacy}
  \[
    \begin{aligned}[t]
      (\phi\circ{T})u 
                &=\sin^{2}\Big(\frac{\pi}{2} - \frac{\pi}{2}|2u-1|\Big) = \sin^{2}(\pi{u})={\Big(2\sin(\frac{\pi}{2})\cos(\frac{\pi}{2})\Big)}^{2}=4\sin^{2}(\frac{\pi}{2}u)\Big(1-\sin^{2}\frac{\pi}{2}u\Big) \\
                                    &=({S}\circ\phi)u.
    \end{aligned}
\]
And it follows that $\phi$ is a conjugacy between the two systems.
\end{Example}

### 2022-01-18 projectReport/chapter3.tex
Worked through all tex errors

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) by examining  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

However, the actual state $u$ of a system is seldom, if ever, fully known. the most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: \ednote{B: Perhaps this edit makes more sense? M: Maybe replace difficulty by saying that we need to consider two aspects. }
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

@@

We define first the concepts of homeomorphism and embedding:
\begin{Definition}\rm
  [\bf {Homeomorphism}]\label{Dfn_homeo}\rm
  A homeomorphism is a function $f:Z\rightarrow Y$ between two topological spaces $Z$ and $Y$ that is continuous, bijective and has a continuous inverse. 
\end{Definition}

\begin{Definition}
  [\bf {Embedding}]\label{Dfn_embed}\rm
  Consider a homeomorphism $f:Z\rightarrow Y$ for $Y\subset X$. $Z$ is said to be embedded in $X$ by $f$.
\end{Definition}

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from \cite{takens1981detecting}.

\begin{Theorem} 
	[\bf Takens Embedding Theorem (adopted from \cite{takens1981detecting}] \label{Thm_Takens}
         Let $W$ be a compact manifold of dimension $m$, and $d\ge m$ so that $2d$ is an integer. It is a 
            generic property for the pair $(T, \theta)$,  where $T:W \to W$ is
            a smooth diffeomorphism, and $\theta:W \to \mathbb{R}$ a smooth function, the map $\Phi_{2d,\theta}:W \to \mathbb{R}^{2d+1}$ defined on $W$ by 
            $\Phi_{2d,\theta}(w) := (\theta(T^{-2d}w)\ldots,\theta(T^{-1}w),\theta(w))$
            is a diffeomorphic embedding; by `smooth' we mean at least $C^2$. Consequently, there exists a map $F_\theta: \Phi_{2d,\theta}(W) \to \Phi_{2d,\theta}(W)$ defined by $$F_\theta: (\theta(T^{-2d}w),\ldots,\theta(T^{-1}w),\theta(w)) \mapsto 
            (\theta(T^{-2d+1}w),\ldots,\theta(w),\theta(Tw))$$
           so that $(W,T)$ is topologically conjugate to 
            $(\Phi_{2d,\theta}(W), F_\theta)$.    
\end{Theorem}

@@

Below we reproduce Figure 1 from \cite{Supp}.

@@

\section{Practical Limitations to Takens' Theorem}
Takens' Embedding Theorem is a powerful result, and it provides compelling reason to believe that one could conceivably reconstruct a system conjugate accurately to the underlying system. 
Nonetheless, it does present some severe practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$ , our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

@@

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. (Cite) \ednote{B: I remember you saying this in one of our sessions, but don't have a source for it. Do you have a specific reference?}

>>>>

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a `nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) by examining  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

However, the actual state $u$ of a system is seldom, if ever, fully known. The most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: \ednote{B: Perhaps this edit makes more sense? M: Maybe replace difficulty by saying that we need to consider two aspects. }
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

@@

We define first the concepts of homeomorphism and embedding:
\begin{Definition}\rm
  {\bf(Homeomorphism)}\label{Dfn_homeo}\rm
  A homeomorphism is a function $f:Z\rightarrow Y$ between two topological spaces $Z$ and $Y$ that is continuous, bijective and has a continuous inverse. 
\end{Definition}

\begin{Definition}
  [\bf{Embedding}]\label{Dfn_embed}\rm
  Consider a homeomorphism $f:Z\rightarrow Y$ for $Y\subset X$. $Z$ is said to be embedded in $X$ by $f$.
\end{Definition}

Takens Theorem states a result establishing a relationship between the observed and underlying dynamical systems by showing that the concatenation of a sufficiently large number of previous observations into a vector will, under certain conditions, generate a map between the vectors from the respective systems.  We formulate the theorem from~\cite{takens1981detecting}.

\begin{Theorem} 
	[\bf Takens Embedding Theorem (adopted from~\cite{takens1981detecting})]\label{Thm_Takens}
         Let $W$ be a compact manifold of dimension $m$, and $d\ge m$ so that $2d$ is an integer. It is a 
            generic property for the pair $(T, \theta)$,  where $T:W \to W$ is
            a smooth diffeomorphism, and $\theta:W \to \mathbb{R}$ a smooth function, the map $\Phi_{2d,\theta}:W \to \mathbb{R}^{2d+1}$ defined on $W$ by 
            $\Phi_{2d,\theta}(w) := (\theta(T^{-2d}w)\ldots,\theta(T^{-1}w),\theta(w))$
            is a diffeomorphic embedding; by `smooth' we mean at least $C^2$. Consequently, there exists a map $F_\theta: \Phi_{2d,\theta}(W) \to \Phi_{2d,\theta}(W)$ defined by \[F_\theta: (\theta(T^{-2d}w),\ldots,\theta(T^{-1}w),\theta(w)) \mapsto 
            (\theta(T^{-2d+1}w),\ldots,\theta(w),\theta(Tw)) \]
           so that $(W,T)$ is topologically conjugate to 
            $(\Phi_{2d,\theta}(W), F_\theta)$.    
\end{Theorem}

@@

Below we reproduce Figure 1 from~\cite{Supp}.

@@

\section{Practical Limitations to Takens' Theorem}
Takens' Embedding Theorem is a powerful result, and it provides compelling reason to believe that one could conceivably reconstruct a system conjugate accurately to the underlying system. 
Nonetheless, it does present some severe practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$, our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

@@

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. (Cite) \ednote{B:I remember you saying this in one of our sessions, but don't have a source for it. Do you have a specific reference?}
